Follow-up Samples

### 2017-03-06 submission/manuscript_outline_20161024.Rmd
Background section changes to active voice

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years the rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].

Over the last few years studies have shown how the bacterial microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  There has also been research into how the bacterial microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  These studies have helped to provide a tantalizing link between the bacterial microbiome and CRC.  Building off of these findings, there has been promising work on the bacterial microbiome and its ability to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies suggest that the bacterial microbiome might change after treatment there remains limited information on the bacterial microbiome response to lesion (adenoma and carcinoma) removal.

In this study we tested the hypothesis that the bacterial microbiome changes between pre (initial) and post (follow up) removal of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We utilized Random Forest to build two models.  The first model was built to classify lesion versus non-lesion (normal) samples while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models, as well as specific OTUs within them, performed pre and post lesion removal helped inform us as to whether initial and follow up samples were changing and whether it was towards a more normal bacterial microbiome or not.  We also investigated the two models for similar important OTUs to identify the crucial OTUs for not only classifying lesion or normal samples but also initial versus follow up.

>>>>

Colorectal cancer (CRC) is currently the third most common cause of cancer deaths [@jemal_cancer_2010; @haggar_colorectal_2009].  The rate of disease mortality has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease per year [@haggar_colorectal_2009].

Recent studies in humans have shown that both the bacterial community and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].  Further, Dejea, et al. observed that bacterial communities are altered between normal and tumor tissue [@dejea_microbiota_2014].  Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.  Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.  Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies show how our gut bacteria can impact CRC progression via a changed community or invations by more inflammatory bacteria [@flynn_metabolic_2016].  They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).

In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.  This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.
